### 2019-11-01 paper/paper.tex
Finish updating citations

Spatial patterns in neural connectivity provide clues about the constraints
under which brains evolve and develop. Whether or not patterning governs
information processing, models of the development of information processing
circuits can be validated in terms of their ability to account for biological
patterning (\citealp{purves_iterated_1992}; see also
\citealp{wilson_what_2015,bednar_cortical_2016}). One of the most distinctive patterns of
neural connectivity is found in the rodent barrel cortex
(\citealp{woolsey_pattern_1948,woolsey_structural_1970,welker_structure_1974}, see
\citealp{Fox2008}). Understanding how this particular pattern forms might shed
light on the interplay of genetic and extrinsic factors that shape cortical
circuits, and on a wealth of anatomical, physiological, and behavioural data
obtained using this model system.

The barrels are large cortical columns, identifiable from birth in layer 4 of
rodent primary somatosensory cortex as dense clusters of thalamocortical
axons, which are enclosed by borders a few neurons thick from postnatal day 3
\citep{Erzurumlu2012}. In the plane tangential to the cortical surface the
barrels constitute a map of the arrangement of the whiskers on the face, such
that the cells of adjacent barrels respond most strongly and most quickly to
deflection of adjacent whiskers \citep{Armstrong-James1992}. The barrels
reflect subcortical whisker maps comprising cellular aggregates known as
barrelettes in the brainstem and barreloids in the thalamus
\citep{Killackey1980}.

Each whisker and its corresponding barrelette, barreloid, and barrel can be
identified by its position on a two-dimensional grid
\citep{Zucker1969,Killackey1980,Loos1976}. The barrels are normally up to
twice as long in the axis aligned to the grid ‘rows’ compared to the ‘arcs’,
and the rows are visible from day 2 \citep{Rebsam2002} whereas the arcs become
distinct by day 3 \citep{Erzurumlu2012,Rebsam2002}. In metabotropic glutamate
receptor 5 (mGlu5) knockouts the barrel rows develop but the arcs do not
(\citealp{Hannan2001} see also \citealp{Fox2008}), thus distinct processes
likely guide development of the barrels with respect to the orthogonal grid
axes.

Barrel formation requires afferent input, driven by whisker stimulation and by
thalamic calcium waves that propagate via mGlu5-receptor activation
(\citealp{Anton2019}, see also \citealp{Anton2018}). Patterning also depends
on a complex network of axon guidance processes, which includes interactions
between guidance molecules like ephrin-A5 and A7 \citep{Miller2006} and
between adhesion molecules like cadherin-6 and 8 \citep{greig_molecular_2013,Bishop2000},
and the binding of p75 to neurotropins (\citealp{Shimogori2005,Bishop2002};
see also \citealp{Dye2011a,Dye2011b}). This network is orchestrated by
interactions between morphogens Fgf8 and Fgf17 and transcription factors Emx2,
Pax6, Sp8, and Coup-tf1 \citep{Shimogori2005,Bishop2000}, which are expressed
in gradients that mark orthogonal axes and can be manipulated to cause the
barrels to stretch, shrink, shift, skew, and even duplicate
\citep{Shimogori2005,Assimacopoulos2012,Borello2014,Sahara2007,Ypsilanti2016,Sur2005}.

The boundaries of the barrels form polygonal shapes that together create a
Voronoi tesselation across the cortical sheet, wherein all points enclosed by
a given boundary are closer to its center than to the center of any other
boundary \citep{Senft1991}. This patterning suggests that the barrel topology
may be preserved in the initial projection of the thalamocortical axons into
the cortex, and that a barrel forms by the lateral branching of the axons from
their initial centres of mass, which ceases upon contact with the axons
branching from adjacent centres. According to this model the barrel pattern
could be entirely determined by the distribution of the initial centre points,
which would itself be determined by guidance cues. However, the assumption of
pre-arranged centres is difficult to resolve with the observation that
thalamocortical axons arrive in the cortical plate as an undifferentiated
bundle prior to barreloid formation (see \citealp{Erzurumlu2012} for a
review).

@@

\cite{Karbowski2004} developed what remains, to the best of our knowledge, the
only dynamical model of cortical arealization capable of generating
predictions about the influence of signalling molecules on the size, shape,
and topological organization of emergent cortical fields. This model describes
competition between thalamocortical projections for territory along a
one-dimensional string of cortical cells, e.g., along an anterior-posterior
axis. The equations of the model define how the density of connections, $c$,
and the density of axon branches, $a$, interact at time $t$ at each cortical
location $x$, for $N$ thalamocortical projections indexed by $i$. The Voronoi
tesselation used to describe barrel patterning is a two-dimensional structure,
hence we first extended the original model for simulation on a two-dimensional
cortical sheet, and describe it here as such (i.e., substituting $x$ with the
two-dimensional position vector $\mb{x}$ and substituting terms of the form
$\frac{d}{dx}$ with the two-dimensional spatial gradient operator $\nabla$).

@@

We first verified that the system given by Eqns. \ref{eq:dc} \&
\ref{eq:da} described a two dimensional extension of the axon
branching model in \citet{Karbowski2004}. We showed that in the presence of
three guidance gradients varying only in one dimension (along the long axis of
an elliptical domain), five thalamocortical axon types, each interacting
uniquely with the three guidance gradients, would segregate into five separate
regions. We investigated a number of configurations of the guidance gradients,
varying their shape, $\rho(\mb{x})$, and number, $M$, and their effect on a
varying number of thalamocortical projections, $N$. In general, we were not
able to produce structure in the resulting patterns of the connections,
$c(\mb{x},t\rightarrow\infty)$, which was not present in the guidance
patterns.

Complex structure can indeed be observed in protein expression for some genes,
such as the molecules Bhlhb5 \citep{Joshi2008} and Lmo4
\citep{Cederquist2013,greig_molecular_2013}. These expression patterns can therefore
provide a theoretical mechanism for the guidance of axons into distinct
regions. The question of how such structures are generated becomes one of of
how gene transcriptional interactions and mRNA diffusion and movement achieves
this (\cmnt{Fixme: find a ref on structure in gene transcription}). However,
complex and stable structure in gene expression data is commonly reported;
monotonically varying expression patterns, such as those for Pax6, Emx2, Sp8
and Coup-tf1 which display graded expression across the entire isocortex
(\cmnt{REFS}) are more typical. Furthermore, the complex structure observed in
the whisker barrel field has not, to our knowledge, been linked to a complex
set of gene expression patterns matching the scale of the barrels. To
investigate how the whisker barrel structure might form in the presence of
only monotonically varying morphogen gradients, we turned our attention to the
competition for connections built into the model.

@@

% Include figure ../boundaries/barreloids_vanderloos_graph.png
\begin{figure*}
\begin{center}
\includegraphics[width=0.7\textwidth]{../boundaries/barreloids_vanderloos_graph.png}
\end{center}
%\vspace{-0em}
\caption{Barreloid centres taken from \cite{van_der_loos_1976}, Fig.~2
  are shown as black circles. It is assumed that there are at least
  two linearly increasing, orthogonal signalling gradients in the
  thalamus (side graphs - addme!), such that the location of each
  barreloid within the graded thalamic signal field can differentiate
  the efferent axons, with the result that axons from each barreloid
  interact to varying degrees with cortical signalling
  gradients. Coloured arrows show, for each barreloid, the magnitude
  of the interaction that its thalamocortical projection will have
  with the four signalling gradients in the cortex. The mappings from
  barreloid location to interaction strength are linear.}
\label{fig:barreloids}
\end{figure*}

>>>>

Spatial patterns in neural connectivity provide clues about the constraints
under which brains evolve and develop. Whether or not patterning governs
information processing, models of the development of information processing
circuits can be validated in terms of their ability to account for biological
patterning (\citealp{purves_iterated_1992}; see also
\citealp{wilson_what_2015,bednar_cortical_2016}). One of the most distinctive patterns of
neural connectivity is found in the rodent barrel cortex
(\citealp{woolsey_pattern_1948,woolsey_structural_1970,welker_structure_1974}, see
\citealp{fox_barrel_2008}). Understanding how this particular pattern forms might shed
light on the interplay of genetic and extrinsic factors that shape cortical
circuits, and on a wealth of anatomical, physiological, and behavioural data
obtained using this model system.

The barrels are large cortical columns, identifiable from birth in layer 4 of
rodent primary somatosensory cortex as dense clusters of thalamocortical
axons, which are enclosed by borders a few neurons thick from postnatal day 3
\citep{erzurumlu_development_2012}. In the plane tangential to the cortical surface the
barrels constitute a map of the arrangement of the whiskers on the face, such
that the cells of adjacent barrels respond most strongly and most quickly to
deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. The barrels
reflect subcortical whisker maps comprising cellular aggregates known as
barrelettes in the brainstem and barreloids in the thalamus
\citep{killackey_pattern_1980}.

Each whisker and its corresponding barrelette, barreloid, and barrel can be
identified by its position on a two-dimensional grid
\citep{zucker_coding_1969,killackey_pattern_1980,van_der_loos_barreloids_1976}. The barrels are normally up to
twice as long in the axis aligned to the grid ‘rows’ compared to the ‘arcs’,
and the rows are visible from day 2 \citep{rebsam_refinement_2002} whereas the arcs become
distinct by day 3 \citep{erzurumlu_development_2012,rebsam_refinement_2002}. In metabotropic glutamate
receptor 5 (mGlu5) knockouts the barrel rows develop but the arcs do not
(\citealp{hannan_plc-1_2001} see also \citealp{fox_barrel_2008}), thus distinct processes
likely guide development of the barrels with respect to the orthogonal grid
axes.

Barrel formation requires afferent input, driven by whisker stimulation and by
thalamic calcium waves that propagate via mGlu5-receptor activation
(\citealp{anton-bolanos_prenatal_2019}, see also \citealp{anton-bolanos_developmental_2018}). Patterning also depends
on a complex network of axon guidance processes, which includes interactions
between guidance molecules like ephrin-A5 and A7 \citep{miller_epha7-ephrin-a5_2006} and
between adhesion molecules like cadherin-6 and 8 \citep{greig_molecular_2013,bishop_regulation_2000},
and the binding of p75 to neurotropins (\citealp{shimogori_fibroblast_2005,bishop_distinct_2002};
see also \citealp{dye_lifespan_2011,dye_lifespan_2011-1}). This network is orchestrated by
interactions between morphogens Fgf8 and Fgf17 and transcription factors Emx2,
Pax6, Sp8, and Coup-tf1 \citep{shimogori_fibroblast_2005,bishop_regulation_2000}, which are expressed
in gradients that mark orthogonal axes and can be manipulated to cause the
barrels to stretch, shrink, shift, skew, and even duplicate
\citep{shimogori_fibroblast_2005,assimacopoulos_fibroblast_2012,borello_sp8_2014,sahara_sp8_2007,ypsilanti_transcriptional_2016,sur_patterning_2005}.

The boundaries of the barrels form polygonal shapes that together
create a Voronoi tesselation across the cortical sheet, wherein all
points enclosed by a given boundary are closer to its center than to
the center of any other boundary \citep{senft_mouse_1991}. This
patterning suggests that the barrel topology may be preserved in the
initial projection of the thalamocortical axons into the cortex, and
that a barrel forms by the lateral branching of the axons from their
initial centres of mass, which ceases upon contact with the axons
branching from adjacent centres. According to this model the barrel
pattern could be entirely determined by the distribution of the
initial centre points, which would itself be determined by guidance
cues. However, the assumption of pre-arranged centres is difficult to
resolve with the observation that thalamocortical axons arrive in the
cortical plate as an undifferentiated bundle prior to barreloid
formation (see \citealp{erzurumlu_development_2012} for a review).

@@

\cite{karbowski_model_2004} developed what remains, to the best of our knowledge, the
only dynamical model of cortical arealization capable of generating
predictions about the influence of signalling molecules on the size, shape,
and topological organization of emergent cortical fields. This model describes
competition between thalamocortical projections for territory along a
one-dimensional string of cortical cells, e.g., along an anterior-posterior
axis. The equations of the model define how the density of connections, $c$,
and the density of axon branches, $a$, interact at time $t$ at each cortical
location $x$, for $N$ thalamocortical projections indexed by $i$. The Voronoi
tesselation used to describe barrel patterning is a two-dimensional structure,
hence we first extended the original model for simulation on a two-dimensional
cortical sheet, and describe it here as such (i.e., substituting $x$ with the
two-dimensional position vector $\mb{x}$ and substituting terms of the form
$\frac{d}{dx}$ with the two-dimensional spatial gradient operator $\nabla$).

@@

We first verified that the system given by Eqns. \ref{eq:dc} \&
\ref{eq:da} described a two dimensional extension of the axon
branching model in \citet{karbowski_model_2004}. We showed that in the presence of
three guidance gradients varying only in one dimension (along the long axis of
an elliptical domain), five thalamocortical axon types, each interacting
uniquely with the three guidance gradients, would segregate into five separate
regions. We investigated a number of configurations of the guidance gradients,
varying their shape, $\rho(\mb{x})$, and number, $M$, and their effect on a
varying number of thalamocortical projections, $N$. In general, we were not
able to produce structure in the resulting patterns of the connections,
$c(\mb{x},t\rightarrow\infty)$, which was not present in the guidance
patterns.

Complex structure can indeed be observed in protein expression for some genes,
such as the molecules Bhlhb5 \citep{joshi_bhlhb5_2008} and Lmo4
\citep{cederquist_lmo4_2013,greig_molecular_2013}. These expression patterns can therefore
provide a theoretical mechanism for the guidance of axons into distinct
regions. The question of how such structures are generated becomes one of of
how gene transcriptional interactions and mRNA diffusion and movement achieves
this (\cmnt{Fixme: find a ref on structure in gene transcription}). However,
complex and stable structure in gene expression data is commonly reported;
monotonically varying expression patterns, such as those for Pax6, Emx2, Sp8
and Coup-tf1 which display graded expression across the entire isocortex
(\cmnt{REFS}) are more typical. Furthermore, the complex structure observed in
the whisker barrel field has not, to our knowledge, been linked to a complex
set of gene expression patterns matching the scale of the barrels. To
investigate how the whisker barrel structure might form in the presence of
only monotonically varying morphogen gradients, we turned our attention to the
competition for connections built into the model.

@@

% Include figure ../boundaries/barreloids_vanderloos_graph.png
\begin{figure*}
\begin{center}
\includegraphics[width=0.7\textwidth]{../boundaries/barreloids_vanderloos_graph.png}
\end{center}
%\vspace{-0em}
\caption{Barreloid centres taken from \cite{van_der_loos_barreloids_1976}, Fig.~2
  are shown as black circles. It is assumed that there are at least
  two linearly increasing, orthogonal signalling gradients in the
  thalamus (side graphs - addme!), such that the location of each
  barreloid within the graded thalamic signal field can differentiate
  the efferent axons, with the result that axons from each barreloid
  interact to varying degrees with cortical signalling
  gradients. Coloured arrows show, for each barreloid, the magnitude
  of the interaction that its thalamocortical projection will have
  with the four signalling gradients in the cortex. The mappings from
  barreloid location to interaction strength are linear.}
\label{fig:barreloids}
\end{figure*}
